{\plant}: A package for modelling forest trait ecology \& evolution

### 2015-05-29 ms.tex
Reworking introduction

\begin{document}
\mstitlepage
\noindent
\parindent=1.5em
\addtolength{\parskip}{.3em}
\doublespacing
\linenumbers
\section{Summary}\label{abstract}
\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\itemsep1pt\parskip0pt\parsep0pt
\item
  Population dynamics in forests are strongly size-structured:
  larger plants shade smaller plants while also expending
  proportionately more energy on building and maintaining woody stems.
  Although the importance of size-
  structure for demography is widely recognised, many mechanistic models
  either omit it entirely, or include only coarse approximations.
\item
  Here we introduce the {\plant} package, an
  extensible framework for modelling trait-based demography, ecology and evolution 
  in vegetation, via coupled differential equations.
  At its core, {\plant} is an
  individual-based mechanistic model where plant physiology and demography is mediated by
  traits. Individuals from multiple species can be grown in isolation,
  in patches of competing plants, or in metapopulations under a
  disturbance regime. Dynamics within patches of competing plants are
  resolved using novel extensions of the ``Escalator Boxcar Train''
  technique. Combined effects of trait-, size- and patch-structured
  dynamics are integrated into population level estimates of
  reproductive fitness.
\item
  {\plant} is an open source R package and is available at
  \href{https://github.com/traitecoevo/plant}{github.com/traitecoevo/plant}.
  While accessed from R, the core routines in {\plant} are written in C++.
  The package provides for alternative physiologies and for
  hyper-parameterisation to capture trade-offs among parameters. A
  detailed test suite is provided to ensure correct behaviour of the code.
\item
  We provide worked examples illustrating {\plant}'s use to study the
  influence of functional traits on growth of individual plants, of
  whole patches and on assembly of ecological communities.
\end{enumerate}

@@

Plant growth and demography are fundamentally size- and trait-
structured, influencing dynamics over time-scales ranging from
instantaneous physiological effects to long-term evolutionary outcomes
\citep{Harper-1977, Westoby-2002}. As an individual increases its leaf
area, it increases its potential to generate photosynthate.  In
contrast, as individuals grow, they must use increasing fractions of
this photosynthetic income to building and maintaining support tissues
rather than to new leaves \citep{Givnish-1988, Enquist-2007} and
reproduction \citep{Thomas-2011}. Consequently rates of growth and
mortality change with individual size \citep{Muller-2006, Ruger-2011, Thomas-2011}.
%
The patterns of changing growth rate vary by plant traits; leaf
construction costs and wood density affect growth rates and rates of
tissue replacement \citep{Wright-2010}, while seed size and height at
maturation determine start and end points of ontogenetic trajectories
\citep{Westoby-2002}
%
Strong feedbacks emerge between individuals within a forest due to
competition for light and other resources, such that the growth rate
of one individual depends on the size and traits of nearby
individuals.  This makes it difficult to predict demographic processes
without modelling both individual growth and the interactions between
individuals.
% TODO[RGF]: link back things about natural selection, successional
% dynamics, self thinning, trait evolution and species coexistance.

Although the importance of size-structure for vegetation function and
demography has been long been recognised
\citep[e.g.,][]{Harper-1977, Shugart-1980, Huston-1987},
most large-scale mechanistic models of plant growth either omit it
entirely, or only include coarse approximations
\citep{Cramer-2001, Sitch-2003, Kelley-2013,Dekauwe-2014}. The likely
reason is the computational
challenge of modelling size-structured dynamics. 
% TODO[RGF]: I don't see the link with the above?  Can this be
% shortened/tightened?  It's too long for a 3k word paper
During the 1980s and
early 1990s, size-structured ``gap'' models were widely used
\citep[e.g.][]{Shugart-1980,Huston-1987}. These gap models were
parameterised in terms of growth using statistical functions; this meant 
that data had to be collected for each species, limiting widespread application. 
The next generation of models focussed on establishing core physiology for 
a limited number of functional types and
sacrificed details of demography in order to achieve global coverage
\citep{Woodward-1995,Haxeltine-1996}. This history has brought us to the
current situation where all but a couple
\citep[e.g.][]{Moorcroft-2001, Smith-2014} of the most popular vegetation
models lack size-structure. Instead, most models
are focused on modelling fluxes carbon and water, but say little about
demographic behaviour; consequently they are also unable to account for
any non-linear demographic feedbacks that may arise.
% While many vegetation models still lack size structure, the importance of
% size-structure for modelling phenomena such as allocation and production is
% being increasingly noted \citep{Moorcroft-2001, Dekauwe-2014, Falster-2011}.

Similarly, most models used in theoretical ecology to explore
questions about species coexistence also lack size-structured dynamics
\citep[e.g.][]{MacArthur-1967,Levin-1974,Tilman-1985,Geritz-1995, 
Calcagno-2006, Leimar-2013}.  It is common to assume
competitive interactions are influenced by size-related traits such as
adult size and seed size, but detailed size-structured demography is
rarely considered, at least for plant models \cite[for animal
examples, see][]{Deroos-1988, Deroos-1992}. Consequently, it has been
difficult to connect mechanisms described in these simplistic models
with type of data often collected in the field, such as physiological
leaf traits and demographic data on growth and mortality in long-term
monitoring plots \citep{Adler-2013}.

In this note, we introduce the {\plant} package for R \citep{R-2015};
a mechanistic framework for studying the effects of size structure and
trait variation on the demography of individual plants, of patches of
competing plants and of meta-populations structured by a prevailing
disturbance regime.
%
The {\plant} package builds on much existing work investigating the
effect of traits on vegetation properties and the mechanisms
facilitating coexistence of trait mixtures
\citep{Kohyama-1993,Deroos-1997,Moorcroft-2001,Falster-2011,Falster-2015}.

% TODO[RGF]: restructure here - I'm getting lost
{\plant} provides a platform for investigating how morphological and
physiological traits influence growth rate and shade tolerance, thus
potentially giving a theoretical foundation for observed empirical
patterns \citep{Baltzer-2007,Wright-2010}.

% NOTE[RGF]: I dropped the appeal to novelty (growing an individual
% plant) because I don't find them very convincing.
While the applications highlighted reflect our own interests in understanding
trait-based demography and community assembly, we expect the methods provided
to be useful in variety of contexts. . Being driven by traits,
the model can be extended to potentially very many species. Users can also 
add their own physiological models. Individual plants
can then be  situated in a patch with competitors and the demography of a
competing population  investigated, thereby capturing phenomena such as multi-
species self-thinning,  successional transitions and age-related changes in
vegetation productivity. {\plant}  models the emergent behaviour of
patches deterministically, based on a continual  size-density distribution. As
such, it is not required to specify a particular patch size.  Finally,
{\plant} allows for emergent phenomena across a metapopulation of patches
to  be investigated, thereby providing a workable mechanism to scale from
individual plant to  vegetation, incorporating the non-linear effects of size
and traits on vegetation outcomes. 
% To strigul?
In  this sense {\plant} is similar to
the \textsc{Ecosystem Demography} \citep{Moorcroft-2001}  and \textsc{LPJ-GUESS}
\citep{Smith-2014} models. However, whereas these other models were designed
to primarily capture large-scale regional biogeochemistry driven by climatic
variables,  {\plant} was designed so that the rules driving the
demography of individual trees, patches and metapopulations could be
manipulated and studied in detail. The three models are therefore highly 
complimentary, they also connect with the increasingly sophisticated statistical 
methods now available to analyse size-structured data
\citep[e.g.][]{Metcalf-2013}.

We describe the general approach of the package and then sections with
details at different levels of ecological organisation. Each section
provides a short description of the required methods and then example 
applications. Detailed technical
documents are provided as supplementary information (see Appendices),
and updated versions of technical documents are also available within
the {\plant} package itself.

@@

% NOTE: Keep the repetition with above under control.

{\plant} is a mechanistic model, meaning the dynamics of the system
arise from rules specifying how individuals grow and interact.  The
core rules in {\plant} are about the short-term physiological
functioning of an individual plant and how these are influenced by its
traits, size and light environment.
%
On top of this physiological model, the package implements methods for
population and adaptive dynamics (Fig.  \ref{fig:schematic}) following
methods described by \citet{Falster-2011} and
\citet{Falster-2015}. 
%
The dynamics at higher levels of organisation arise as emergent
properties, driven by growth physiology, competition for light and
disturbance (Fig.  \ref{fig:schematic}). Demographic phenomena can be
studied at the three levels: individual plants, stands of competing
plants and entire metapopulations.
%
We developed {\plant} as an extended thought experiment for helping to
understand competition and succession by having these driven by
low-level processes rather than being explicitly implemented in the
model.
% NOTE: I think that foreshaddowing some of the competition kernels
% idea here is sensible.
%
% TODO: Somewhere, perhaps above, perhaps highlight the issue of doing
% coexistance modelling with explicit modelling - the models are so
% well understood it's easy to drive the behaviour.  But we want to
% have something where we just get coexistence to study how tradeoffs
% generate it.
%
% TODO: perhaps salavageable?
%
%   {\plant} was developed with this style of analysis in mind, while
%   also allowing for a richer set of ecological dynamics than is
%   possible in population models lacking size structure.

@@

The core of {\plant} is a model for an individual plant's
physiological strategy as specified by its traits
(Fig. \ref{fig:schematic}a). This model estimates rate of biomass
production for a plant, given its size and a light environment. Gross
photosynthetic income is estimated from total leaf area and the light
distribution across the plant's canopy. Costs of tissue respiration
and turnover are then subtracted. The remaining biomass is then
allocated between growth and reproduction.  
%
The key outputs needed by higher levels (see below) is growth rate,
mortality rate and rate of seed production.  However additional
quantities are also computed, such as total assimilation and the
respiration, turnover and allocation rates for different tissues (see
Appendix \ref{sec:plant})

@@

We assume that patches are vertically but not horizontally
structured. In other words, we account for size differences but not spatial 
layout within the patch. Similar assumptions are found in many models
simulating size-structured dynamics
\citep{Kohyama-1993, Huston-1987, Moorcroft-2001, Smith-2014}. In addition, 
we assume that patches are very large -- this means the population dynamics 
within the patch behave deterministically \citep{Deroos-1997} and can therefore 
be approximated via a Partial Differential Equation (\textsc{pde}) (see Appendix
\ref{sec:demography}). The same \textsc{pde} captures average behaviour
across a large number of small patches \citep{Moorcroft-2001}. The model may 
therefore capture average demographic behaviour across a range of patch sizes.

>>>>

\begin{document}
\mstitlepage
\noindent
\parindent=1.5em
\addtolength{\parskip}{.3em}
\doublespacing
\linenumbers
\section{Summary}\label{abstract}
\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\itemsep1pt\parskip0pt\parsep0pt
\item
  Population dynamics in forests are strongly size-structured:
  larger plants shade smaller plants while also expending
  proportionately more energy on building and maintaining woody stems.
  Although the importance of size-
  structure for demography is widely recognised, many mechanistic models
  either omit it entirely, or include only coarse approximations.
\item
  Here we introduce the {\plant} package, an
  extensible framework for modelling size- and trait-based demography, 
  ecology and evolution in vegetations.
  At its core, {\plant} is an
  individual-based mechanistic model where plant physiology and demography is mediated by
  traits. Individuals from multiple species can be grown in isolation,
  in patches of competing plants, or in metapopulations under a
  disturbance regime. Dynamics within patches of competing plants are
  resolved using novel extensions of the ``Escalator Boxcar Train''
  technique. Combined effects of trait-, size- and patch-structured
  dynamics are integrated into metapopulation level estimates of
  reproductive fitness and vegetation structure.
\item
  {\plant} is an open source R package and is available at
  \href{https://github.com/traitecoevo/plant}{github.com/traitecoevo/plant}.
  While accessed from R, the core routines in {\plant} are written in C++.
  The package provides for alternative physiologies and for
  hyper-parameterisation to capture trade-offs among parameters. A
  detailed test suite is provided to ensure correct behaviour of the code.
\item
% todo
  We provide worked examples illustrating {\plant}'s use to study the
  influence of functional traits on growth of individual plants, of
  whole patches and on assembly of ecological communities.
\end{enumerate}

@@

Plant growth and demography are fundamentally size- and trait-
structured, influencing dynamics over time-scales ranging from
instantaneous physiological effects to long-term evolutionary outcomes
\citep{Harper-1977, Westoby-2002}. As an individual increases its leaf
area, it increases its potential to generate photosynthate.  In
contrast, as individuals grow, they must use increasing fractions of
this photosynthetic income to building and maintaining support tissues
rather than to new leaves \citep{Givnish-1988, Enquist-2007} and
reproduction \citep{Thomas-2011}. Consequently rates of growth and
mortality change with individual size \citep{Muller-2006, Ruger-2011, Thomas-2011}.
The patterns of changing growth rate vary by plant traits; leaf
construction costs and wood density affect growth rates and rates of
tissue replacement \citep{Wright-2010}, while seed size and height at
maturation determine start and end points of ontogenetic trajectories
\citep{Westoby-2002}
Strong feedbacks emerge between individuals within a forest due to
competition for light and other resources, such that the growth rate
of one individual depends on the size and traits of nearby
individuals.  This makes it difficult to predict demographic processes
without modelling both individual growth and the interactions between
individuals. The non-linear feedbacks arising from competitive 
interactions are thought to be crucial for understanding phenomena such 
as self thinning, successional transitions, trait evolution and 
species coexistence.

Although the importance of size-structure for vegetation dynamics and diversity, 
has been long been recognised
\citep[e.g.,][]{Harper-1977, Shugart-1980, Huston-1987}, current research in these
areas is dominated by models and theory that either omit size-structure 
entirely, or only include coarse approximations. Theoretical investigations 
exploring questions about niche-based species coexistence and ecological drift 
(neutrality) both primarily rely  on unstructured population models 
\citep[e.g.][]{MacArthur-1967, Tilman-1985, Geritz-1995, Hubbell-2001,
Calcagno-2006, Leimar-2013}. It is common to assume
competitive interactions are influenced by size-related traits such as
adult size and seed size, but detailed size-structured demography is
rarely considered, at least for plant models \cite[for animal
examples, see][]{Deroos-1988, Deroos-1992}. Consequently, it has been
difficult to connect and test the mechanisms described in these simplistic models
with types of data collected in the field, such as physiological
leaf traits or demographic data from long-term monitoring plots \citep{Adler-2013}.
Similarly, most large-scale models of vegetation dynamics do not include
size-structured demography \citep{Cramer-2001, Sitch-2003, Kelley-2013,Dekauwe-2014}. 
During the  1980s, size-structured ``gap'' models were widely used
\citep[e.g.][]{Shugart-1980,Huston-1987}. However, the next generation of models 
focussed more heavily on biogeochemistry at the expense of demography 
\citep[e.g.][]{Woodward-1995,Haxeltine-1996}. While the importance
of size-structure is starting again to be recognised \citep{Moorcroft-2001, 
 Falster-2011, Dekauwe-2014}, all but a couple \citep[e.g.][]{Moorcroft-2001, Smith-2014} 
of the most popular, large-scale vegetation models lack any real demography or 
size-structure.

In this note, we introduce the {\plant} package for R \citep{R-2015};
a mechanistic framework for studying the effects of size structure and
trait variation on the demography of individual plants, of patches of
competing plants and of meta-populations structured by a prevailing
disturbance regime. In the theoretical ecology literature it is common to
analyse the full behaviour of a model under its equilibrium conditions.
{\plant} was developed with this style of analysis in mind, while also 
allowing for a richer set of ecological dynamics than is possible in 
population models lacking size structure. {\plant} is thus part
of an extended thought experiment investigating how species differing
in traits are able to coexist with one another \citep{Falster-2011,Falster-2015}.
Understanding and incorporating the demographic outcomes of trait variation 
is ultimately needed if we are to properly understand how trade-offs in plant function 
enable species coexistence.

At the same time, we expect the {\plant} model to be useful for modelling
other  phenomena, beyond questions about species coexistence. The internal
structure  of the {\plant} model is superficially similar to the
\textsc{Ecosystem Demography}  \citep{Moorcroft-2001}  and \textsc{LPJ-GUESS}
\citep{Smith-2014} models. Like  {\plant}, these models include size-
structured competition within a metapopulation of  "patches". However,
whereas these models were designed to capture large-scale regional
biogeochemistry driven by climatic variables and rely on simulations,
{\plant} was designed so that the rules  driving the dynamics of the system
could be manipulated and emergent demographic  behaviours studied in detail.
In {\plant} it is possible to grow plants in isolation, in patches of
competing plants and in disturbance-driven metapopulations. The package is
thus designed to sit in the middle ground between a large-scale coupled
simulation model and simple abstract theoretical device. As examples, we
expect the package will prove  useful for investigating how morphological and
physiological traits influence growth rate and shade tolerance, and emergent
phenomena such as multi-species self-thinning,  successional transitions and
age-related changes in vegetation productivity.

Below we describe the general approach of the package, then provide  sections
with details at different levels of ecological organisation.  Detailed
technical documents are provided as supplementary information  (see
Appendices), and updated versions of technical documents are also available
within the {\plant} package itself.

@@

{\plant} is a mechanistic model, meaning the dynamics of the system
arise from rules specifying how individuals grow and interact.  The
core rules in {\plant} are about the short-term physiological
functioning of an individual plant and how these are influenced by its
traits, size and light environment. Being driven by traits, the model can be extended to potentially 
very many species. Users can also  add their own physiological models.

On top of this physiological model, the package implements methods for
population and adaptive dynamics (Fig.  \ref{fig:schematic}) following methods
described by \citet{Falster-2011} and \citet{Falster-2015}. Demographic
phenomena can be studied at the three levels: individual plants, stands of
competing plants and entire metapopulations. The dynamics at these higher
levels of organisation arise as emergent properties, driven by growth
physiology, competition for light and disturbance (Fig.  \ref{fig:schematic}).
We model the emergent behaviour of patches deterministically, based on a
continual  size-density distribution. As such, it is not required to specify a
particular patch size.

@@

The core of {\plant} is a model for an individual plant's
physiological strategy as specified by its traits
(Fig. \ref{fig:schematic}a). This model estimates rate of biomass
production for a plant, given its size and a light environment. Gross
photosynthetic income is estimated from total leaf area and the light
distribution across the plant's canopy. Costs of tissue respiration
and turnover are then subtracted. The remaining biomass is then
allocated between growth and reproduction.  
The key outputs needed by higher levels (see below) is growth rate,
mortality rate and rate of seed production.  However additional
quantities are also computed, such as total assimilation and the
respiration, turnover and allocation rates for different tissues (see
Appendix \ref{sec:plant})

@@

We assume that patches are vertically but not horizontally
structured. In other words, we account for size differences but not spatial 
layout within the patch. Similar assumptions are found in many models
simulating size-structured dynamics
\citep{Shugart-1980,Kohyama-1993, Huston-1987, Moorcroft-2001, Smith-2014}. In addition, 
we assume that patches are very large -- this means the population dynamics 
within the patch behave deterministically \citep{Deroos-1997} and can therefore 
be approximated via a Partial Differential Equation (\textsc{pde}) (see Appendix
\ref{sec:demography}). The same \textsc{pde} captures average behaviour
across a large number of small patches \citep{Moorcroft-2001}. The model may 
therefore capture average demographic behaviour across a range of patch sizes.
